Information in recombination junctions can both compress and improve inferred tree sequences

### 2024-01-04 paper/writeup.tex
added discrepancy concept figure

\subsection{Extend Edges}}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying short paths (2 edges) in a tree.
    If neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
    suppose there exists a path which contains a node $3$,
    with parent and child nodes $4$ and $0$, respectively.
    If there is an edge between the nodes $4$ and $0$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
    This action reduces the length of $4 \to 0$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such ancestral paths 
    over the entire tree sequence.
    With this algorithm \ref{alg:edge} we only extend existing edges
    over a larger interval on the genome,
    and remove unnecessary edges in the process.

@@

\par Now with an algorithm in place
 we need a metric to discern
 if trees modified with edge extend
 in fact correctly infer haplotypes in our original tree sequence.
 One such function a so called \textit{discrepancy function}, can be called on a pair of tree 
 To count all the discrepancies we must begin by counting differences for each node in the tree sequence.

\begin{figure}

    \caption{
        Conceptual figure displaying what discrepancy is?
        TODO: brainstorm.
        \label{fig:conceptual_discrepancy}
    }
\end{figure}

>>>>

\subsection{Extend Edges}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction? (reduces edges by 1/3?)
    We define a \textit{tree sequence}, denoted $\T$, 
    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
     each $T_k$ is a tree describing genealogical relationships between samples
     over a distinct portion of the genome $[a_{k-1}, a_k)$.
    Given any tree sequence, our goal is to
    identify areas of implied inheritance of haplotypes.
    We do so by identifying short paths (2 edges) in a tree.
    If neighboring trees contain edges with 
    identical ancestor (parent) and descendant (child) nodes, 
    we assume that there is implied coalescence of haplotypes between nodes, 
    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
    suppose there exists a path which contains a node $3$,
    with parent and child nodes $4$ and $0$, respectively.
    If there is an edge between the nodes $4$ and $0$ in trees 
    $T_{k+1}$ (or $T_{k-1}$),
    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
    This action reduces the length of $4 \to 0$ on the genome,
    and in some cases, completely removes the edge. 
    We now perform this action on all such ancestral paths 
    over the entire tree sequence.
    With this algorithm \ref{alg:edge} we only extend existing edges
    over a larger interval on the genome,
    and remove unnecessary edges in the process.

@@

\par Now with an algorithm in place
 we need a metric to discern
 if trees modified with edge extend
 in fact correctly infer haplotypes in our original tree sequence.
 One such function a so called \textit{discrepancy function}, can be called on a pair of tree sequences.
 To count all the discrepancies we must begin by counting differences for each node in the tree sequence.

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        Conceptual figure displaying what discrepancy is?
        TODO: brainstorm.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}
